### 2018-01-10 submission/manuscript.Rmd
rewrite background - add clearer picture of the confusion in current understanding

@@

This intense focus on identifying biomarkers has found strong associations with resident mouth microbes as potential CRC-associated microbes in the gastrointestinal (GI) tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014]. The main bacteria of interest, arising from this set of microbes, has been those within the *Fusobacterium* genus. Yet, the question remains as to whether or not this is indeed the most important genera to be focusing on, since many microbiota-based studies typically have identified a collection of oral microbes rather than single species from a single genera [@baxter_microbiota-based_2016; @zeller_potential_2014]. Based on this discrepancy, the second question we can answer with this meta-analysis is if there is one dominant CRC-associated genera that can be identified across studies.

The identification of microbial biomarkers has mostly dominated the study of CRC and has had an unattended consequence of reducing the focus on changes that occur within the underlying resident community.  This has been borne out by the majority of previous studies, within stool, tissue, and the only meta-analysis investigating this area to date, that focus predominately on biomarker identification. Yet, how these CRC-associated microbes interact with their community may be an important component to consider, and has not been investigated in great detail by other meta-analyses. In response to this gap, our study aims to answer whether there are consistent detectable community differences as disease severity increases.

In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, examines differences between stool and tissue microbiota in the context of CRC, and takes a more community centric approach rather than a biomarker focused approach to investigating commonalities across study for the microbiota and CRC. Importantly, this community centric approach could provide valuable insights into the importance of accounting for the community in CRC disease not previously provided by earlier meta-analysis studies [@shah_leveraging_2017].

>>>>

Although there has been an intense focus on microbiota based biomarker discovery for CRC, the candidate genera seem to be numerous. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in some of these studies mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not necessarily limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, *Escherichia/Shigella* also has good *in vivo* evidence that it could be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential additional markers of CRC with *Bacteroides fragilis* having good mechanistic studies into how this may occur [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. Still, other studies have identified genera like *Roseburia* as increased in CRC while others have reported it to be decreased or have found no difference in any genera at all [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

Most of these studies have focused on carcinoma but the adenoma results are not any clearer at identifying candidate genera. Some groups have focused on more broad scale community metrics and have found that metrics such as richness had decreased in adenoma versus controls while others have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017]. A recent meta-analysis confirmed the associations of mouth-associated genera and *Akkermansia muciniphila* in carcinoma [@shah_leveraging_2017]. However, it only analyzed a total of 509 stool samples and with some recent studies, that were not included, matching that sample size alone it is hard to know how extrapolatable these findings truly are. This meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) that need to be investigated further since a number of these genera have been found to be enriched in controls and not lesion (adenoma or carcinoma) [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].

@@

In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of CRC. Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in CRC and whether broad bacterial community measurements can account for these changes that were not previously provided by earlier meta-analysis studies [@shah_leveraging_2017].
